---
title: "CWS- Offshore Bat Monitoring Project"
subtitle: "Manual Verification Data Summary "
format:
  html:
    grid:
      margin-width: 300px
    title-block-banner: "#1d4566"
    title-block-banner-color: "#F4F4F5"
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-expand: true
    toc-location: left
    css: styles.css
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    colorlinks: true
    geometry:
      - margin=1in
    papersize: letter
date: last-modified
author: 
  - name: "Camila Hurtado"
    affiliation: "Lead, Biodiversity Pathways Ltd."
editor: visual
bibliography: references.bib
nocite: '@*'
---

```{r}
#| label: setup - load libraries and data
#| include: false
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(ggplot2)
library(dplyr)
library(tinytex)
library(quarto)
library(lubridate)
library(flextable)
library(tidyverse)
library(kableExtra)
library(readr)
library(leaflet)
library(sf)

df <- read.csv("Data/species_summary_by_site_month_year.csv")

```

```{r}
#| label: setup - create summary of processing table
#| include: false
#| echo: false
#| eval: true
#| warning: false
#| message: false

#Upload the filerenaming csv
rename <- read.csv("Data/MNA_File_Renaming.csv")

#first identify the specific sites from the filenames
data <- df %>%
  mutate(site = str_extract(filename, "^[^_]+"))

# Summarize which species were found at each site adn by each detector
detailed_summary <- data %>%
  group_by(site) %>%
  summarise(
    n_files = n(),
    kaleidoscope_species = {
      species <- `WA.Kaleidoscope.Auto.ID` %>%
        na.omit() %>%
        str_split("[, ]+") %>%                # Split by comma, space, or both
        unlist() %>%
        .[. != "" & !str_detect(., "(?i)^noise$")] %>%  # Remove empty and Noise only
        unique() %>%
        sort()
      if(length(species) > 0) paste(species, collapse = ", ") else "None"
    },
    sonobat_species = {
      species <- `SB.Species.Auto.ID.verbose` %>%
        na.omit() %>%
        str_split("[, ]+") %>%                # Split by comma, space, or both
        unlist() %>%
        .[. != "" & !str_detect(., "(?i)^noise$")] %>%  # Remove empty and Noise only
        unique() %>%
        sort()
      if(length(species) > 0) paste(species, collapse = ", ") else "None"
    },
    manual_species = {
      species <- `Species.Manual.ID` %>%
        na.omit() %>%
        str_split("[, ]+") %>%                # Split by comma, space, or both
        unlist() %>%
        .[. != "" & !str_detect(., "(?i)^noise$")] %>%  # Remove empty and Noise only
        unique() %>%
        sort()
      if(length(species) > 0) paste(species, collapse = ", ") else "None"
    }
  )


```

```{r}
#| label: setup - heat maps of species detections
#| include: false
#| echo: false
#| eval: true
#| warning: false
#| message: false

# First, make sure your data has the site column
data <- df %>%
  mutate(site = str_extract(filename, "^[^_]+"))

# Corrected function
create_species_matrix <- function(data, species_column, location_column = "site") {
  # Create a dataframe with location and species strings
  species_data <- data %>%
    select(location = all_of(location_column), species_string = all_of(species_column)) %>%
    filter(!is.na(species_string) & species_string != "" & species_string != "None") %>%
    distinct()
  
  # Split comma-separated species into individual rows
  species_long <- species_data %>%
    separate_rows(species_string, sep = ",\\s*") %>%
    filter(species_string != "" & 
           species_string != "None" & 
           species_string != "NoID" & 
           species_string != "NOISE"&
           species_string != "Noise" &
           species_string != "HiLo" ) %>%   
    mutate(present = 1) %>%
    distinct(location, species_string, .keep_all = TRUE)
  
  # Create presence/absence matrix
  species_matrix <- species_long %>%
    pivot_wider(
      names_from = species_string,
      values_from = present,
      values_fill = 0
    ) %>%
    arrange(location)
  
  return(species_matrix)
}

# ============================================================================
# Manual ID
# ============================================================================
manual_matrix <- create_species_matrix(data, "Species.Manual.ID")

# Convert to long format for ggplot
manual_long <- manual_matrix %>%
  pivot_longer(
    cols = -location,
    names_to = "species",
    values_to = "present"
  ) %>%
  mutate(species = factor(species))

manual_long$location <- factor(manual_long$location, 
                               levels = sort(unique(manual_long$location)))

# Custom species ordering: alphabetical, but HighF, LowF, 40K at end
species_list <- unique(manual_long$species)
freq_categories <- c("40K", "HiF", "HiLo", "LoF", "HighF", "LowF")
regular_species <- sort(setdiff(species_list, freq_categories))
ordered_species <- c(regular_species, sort(freq_categories))

manual_long$species <- factor(manual_long$species, levels = ordered_species)

# Create heat map for Manual ID
MV <- ggplot(manual_long, aes(x = species, y = location, fill = factor(present))) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_manual(
    values = c("0" = "gray95", "1" = "#2C7BB6"),
    labels = c("Absent", "Present"),
    name = "Detection"
  ) +
  labs(
    #subtitle = "Species confirmed through manual verification of recordings",
    x = "Species Code",
    y = "Location"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11, color = "gray40"),
    panel.grid = element_blank(),
    legend.position = "bottom"
  )

# ============================================================================
# KALEIDOSCOPE AUTO ID
# ============================================================================

KS_matrix <- create_species_matrix(data, "WA.Kaleidoscope.Auto.ID")

# Convert to long format for ggplot
KS_long <- KS_matrix %>%
  pivot_longer(
    cols = -location,
    names_to = "species",
    values_to = "present"
  ) %>%
  mutate(species = factor(species))

# Truncate species codes to 4 letters for Kaleidoscope
KS_long <- KS_long %>%
  mutate(species = ifelse(
    nchar(as.character(species)) > 4,
    paste0(str_sub(as.character(species), 1, 2), 
           str_sub(as.character(species), 4, 5)),
    as.character(species)
  ))

#organize based on alphabetical order of locations
KS_long$location <- factor(KS_long$location, 
                           levels = sort(unique(KS_long$location)))

# Custom species ordering: alphabetical, but HighF, LowF, 40K at end
KS_species_list <- unique(KS_long$species)
freq_categories <- c("40K", "HiF", "HiLo", "LoF", "HighF", "LowF")
KS_regular_species <- sort(setdiff(KS_species_list, freq_categories))
KS_ordered_species <- c(KS_regular_species, sort(freq_categories))

KS_long$species <- factor(KS_long$species, levels = KS_ordered_species)

# Create heat map for Kaleidoscope
KS_plot <- ggplot(KS_long, aes(x = species, y = location, fill = factor(present))) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_manual(
    values = c("0" = "gray95", "1" = "#2C7BB6"),
    labels = c("Absent", "Present"),
    name = "Detection"
  ) +
  labs(
    #subtitle = "Species identified by Kaleidoscope",
    x = "Species Code",
    y = "Location"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11, color = "gray40"),
    panel.grid = element_blank(),
    legend.position = "bottom"
  )

# ============================================================================
# SONOBAT AUTO ID
# ============================================================================

SB_matrix <- create_species_matrix(data, "SB.Species.Auto.ID.verbose")

# Convert to long format for ggplot
SB_long <- SB_matrix %>%
  pivot_longer(
    cols = -location,
    names_to = "species",
    values_to = "present"
  ) %>%
  mutate(species = factor(species))

SB_long$location <- factor(SB_long$location, 
                           levels = sort(unique(SB_long$location)))

# Custom species ordering: alphabetical, but HighF, LowF, 40K at end
SB_species_list <- unique(SB_long$species)
SB_regular_species <- sort(setdiff(SB_species_list, freq_categories))
SB_ordered_species <- c(SB_regular_species, sort(freq_categories))

SB_long$species <- factor(SB_long$species, levels = SB_ordered_species)

# Create heat map for SonoBat
SB_plot <- ggplot(SB_long, aes(x = species, y = location, fill = factor(present))) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_manual(
    values = c("0" = "gray95", "1" = "#2C7BB6"),
    labels = c("Absent", "Present"),
    name = "Detection"
  ) +
  labs(
    #subtitle = "Species identified by SonoBat",
    x = "Species Code",
    y = "Location"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11, color = "gray40"),
    panel.grid = element_blank(),
    legend.position = "bottom"
  )

```

```{r}
#| label: setup - bat spp definitions
#| include: false
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Create species reference table from scratch
unique_bat_ids <- data.frame(
  Code = c("EPFU", "EPFULACI", "EPFULANO", "LABO", "LABOMYLU", "LACI", 
           "LACILANO", "LANO", "MYCI","MYCIMYLU", "MYCIMYVO","MYEV", "MYLU", "MYLUMYSE", "MYSE", "MYVO", "NOID", "NOISE",
           "40KMYO", "HighF", "LowF"),
  stringsAsFactors = FALSE
)

# Add definitions for each manual ID
unique_bat_ids$Definition <- c(
  "Calls that have diagnostic features identifying it as Eptesicus fuscus",
  "Calls that could be attributed to either Eptesicus fuscus or Lasiurus cinereus",
  "Calls that could be attributed to either Eptesicus fuscus or Lasyonicteris noctivagans",
  "Calls that have diagnostic features identifying it as Lasiurus borealis",
  "Calls that could be attributed to either Lasiurus borealis or Myotis lucifugus",
  "Calls that have diagnostic features identifying it as Lasiurus cinereus",
  "Calls that could be attributed to either Lasiurus cinereus or Lasyonicteris noctivagans",
  "Calls that have diagnostic features identifying it as Lasyonicteris noctivagans",
  "Calls that have diagnostic features identifying it as Myotis ciliolabrum",
  "Calls that could be attributed to either Myotis ciliolabrum or Myotis lucifugus",
  "Calls that could be attributed to either Myotis ciliolabrum or Myotis volans",
  "Calls that have diagnostic features identifying it as Myotis evotis",
  "Calls that have diagnostic features identifying it as Myotis lucifugus",
  "Calls that could be attributed to either Myotis lucifugus or Myotis septentrionalis",
  "Calls that have diagnostic features identifying it as Myotis septentrionalis",
  "Calls that have diagnostic features identifying it as Myotis volans",
  "Bat calls but no grouping category applies",
  "No bat recorded",
  "Various species of Myotis that have a characteristic frequency in the range of 35-40kHz",
  "Various species with pulses having a characteristic frequency higher than ~35kHz",
  "Various species with pulses having a characteristic frequency lower than ~30kHz"
)

# Add Scientific names
unique_bat_ids$ScientificName <- c(
  "Eptesicus fuscus",
  "Eptesicus fuscus / Lasiurus cinereus",
  "Eptesicus fuscus / Lasyonicteris noctivagans",
  "Lasiurus borealis",
  "Lasiurus borealis / Myotis Lucifugus",
  "Lasiurus cinereus",
  "Lasiurus cinereus / Lasyonicteris noctivagans",
  "Lasyonicteris noctivagans",
  "Myotis ciliolabrum",
  "Myotis ciliolabrum / Myotis Lucifugus",
  "Myotis ciliolabrum / Myotis volans",
  "Myotis evotis",
  "Myotis lucifugus",
  "Myotis lucifugus / Myotis septentrionalis",
  "Myotis septentrionalis",
  "Myotis volans",
  "",
  "",
  "",
  "",
  ""
)

# Add common names
unique_bat_ids$CommonName <- c(
  "Big Brown Bat",
  "Big Brown Bat / Hoary Bat",
  "Big Brown Bat / Silver-haired Bat",
  "Eatern Red Bat",
  "Eastern Red Bat / Little Brown Myotis",
  "Hoary Bat",
  "Hoary Bat / Silver-haired Bat",
  "Silver-haired Bat",
  "Western Small-footed Myotis",
  "Western Small-footed Myotis / Little Brown Myotis",
  "Western Small-footed Myotis / Long-legged Myotis",
  "Long-eared Myotis",
  "Little Brown Myotis",
  "Little Brown Myotis / Northern Myotis",
  "Northern Myotis",
  "Long-legged Myotis",
  "Unknown Bat",
  "No Bat",
  "40kHz Frequency Myotis",
  "High Frequency Bat",
  "Low Frequency Bat"
)

# Re-order columns for the table
unique_bat_ids <- unique_bat_ids[, c("CommonName", "ScientificName", "Code", "Definition")]

# If you want just a simple character vector
unique_bat_id_vector <- unique_bat_ids$Code


```

## Land Acknowledgement

Biodiversity Pathways respectfully acknowledges that our work takes place on treaty lands as well as the traditional and unceded territories of First Nations, Inuit, and Métis Peoples across all regions of Canada, whose histories, languages, and cultures are deeply connected to the biodiversity we monitor. We acknowledge the traditional teachings of the lands that we work on, and that reciprocal, meaningful, and respectful relationships with Indigenous peoples make our work possible. We are deeply grateful for their stewardship of these lands, and we are committed to supporting Indigenous-led monitoring programs, while learning Indigenous ways of knowing, being, and doing.

## Background

## Methods

### Data processing

Full-spectrum recordings from the sampling periods were processed using two automatic classifiers: Kaleidoscope's Bats of North America 5.4.0 classifier and Sonobat 3.0's Northeastern North America classifier. Based on documented species ranges, manual identification efforts focused on 6 species: Big Brown Bats (*Eptesicus fuscus*), Silver-haired Bats (*Lasionycteris noctivagans*), Hoary Bat (*Lasiurus cinereus*), Little Brown Myotis (*Myotis lucifugus*), Northen Myotis (*Myotis spententrionalis*) and Tri-colored bat (*Perimyotis subflavus*).

The analysis workflow followed processing standards established by the North American Bat Monitoring Program (NABat) [@reichert2018Guide]. Only recordings that were identified as a bat by Kaleidoscope were manually verified. All recordings with automated classifications underwent complete manual verification. Species identifications were validated using reference call parameters described by @szewczak2018Acoustic, @slough2022New, and @solick2022Bat, in accordance with NABat manual vetting protocols [@reichert2018Guide].

## Results

Following manual verification, hoary bats(*Lasiurus cinereus*), eastern red bats (*Lasirus borealis*), and little brown Myotis (*Myotis lucifugus*) were detected at all surveyed locations ([@fig-MVsummary]). Silver-haired bats (*Lasionycteris noctivagans*) were detected at all sites except IPCA-BH-BAT-07; however, their presence at this site remains plausible given regional distributions. Northern Myotis (*Myotis septentrionalis*) were detected only at the IPCA-STRATH-01 location. As this species is challenging to detect using acoustic methods, non-detection at other sites should not be interpreted as evidence of true absence.

```{r}
#| label: fig-MVsummary
#| fig-cap: "Species confirmed through manual verification across monitoring locations in 2025. Blue tiles indicate species presence; gray tiles indicate absence."
#| echo: false
#| warning: false
#| message: false

if (knitr::is_latex_output()) {
  # PDF output
  print(MV)
} else {
  # HTML output
  print(MV)
}
```

```{r}
#| label: fig-KSsummary
#| fig-cap: "Species detected by Kaleidoscope AutoID across monitoring locations in 2025. Blue tiles indicate species presence; gray tiles indicate absence."
#| echo: false
#| warning: false
#| message: false

if (knitr::is_latex_output()) {
  # PDF output
  print(KS_plot)
} else {
  # HTML output
  print(KS_plot)
}
```

```{r}
#| label: fig-SBsummary
#| fig-cap: "Species detected by Sonobat AutoID across monitoring locations in 2025. Blue tiles indicate species presence; gray tiles indicate absence."
#| echo: false
#| warning: false
#| message: false

if (knitr::is_latex_output()) {
  # PDF output
  print(SB_plot)
} else {
  # HTML output
  print(SB_plot)
}
```

## Recomendations

{{< pagebreak >}}

## Appendix A

Species codes and their definitions

```{r}
#| label: AppendixA
#| echo: false
#| warning: false
#| message: false
if (knitr::is_latex_output()) {
    # PDF output with column wrapping
  unique_bat_ids %>%
    kable(format = "latex", 
          booktabs = TRUE,
          longtable = TRUE) %>%
    kable_styling(
      latex_options = c("striped", "repeat_header"),
      font_size = 9
    ) %>%
    column_spec(1, width = "3.5cm") %>%
    column_spec(2, width = "3.5cm") %>%
    column_spec(3, width = "2cm") %>%
    column_spec(4, width = "6cm")  # This allows wrapping in Definition column
} else {
  # HTML output
  unique_bat_ids %>%
    flextable() %>%
    align(j = 1, align = "left") %>%
    align(j = 2:ncol(unique_bat_ids), align = "center") %>%
    theme_zebra() %>%
    autofit()
}
```

## Literature Cited
